Autonomous Oceanographic Data Collection Estimating Excursion Sets of Vector-valued Gaussian Random Fields

### 2020-06-25 main_new.tex
Updates from Overleaf

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime resource management. Faced with
  limited resources in understanding processes in the water-column,
  the combination of statistics and autonomous systems provide new
  opportunities for experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal ocean processes by providing informative descriptions of
  spatial characteristics of ocean phenomena. Specifically, we define
  a design criterion based on uncertainty in the excursions of
  vector-valued Gaussian random fields, and derive tractable
  expressions for the expected Bernoulli variance reduction in such a
  framework. We demonstrate how this criterion can be used to
  prioritize sampling efforts at locations that are ambiguous, making
  exploration more effective. We use simulations to study the
  properties of methods and to compare them with state-of-the-art
  approaches, followed by results from field deployments with an
  autonomous underwater vehicle as part of a study mapping the
  boundary of a river plume. The results demonstrate the potential of
  combining statistical methods and robotic platforms to effectively
  inform and execute data-driven environmental sampling.

@@

\subsection{Expected Bernoulli Variance on a two dimensional Example}
\label{Sec:UnivarEx}

We illustrate the EBV associated with different designs on a simple
bivariate example. This mimics our river plume application and hence
the first and second component of the random field will be called
\textit{temperature} and \textit{salinity}. We begin
with a \textit{pointwise} example, considering a single bivariate Gaussian distribution (i.e. no spatial elements).

@@

Say we want to study the excursion probability of a bivariate Gaussian
above some threshold, where the thresholds are set equal to the mean;
$\mu_1=t_1=5^o C$ for temperature and $\mu_2=t_2=30$ g/kg for
salinity, and we play with the temperature and salinity correlation
and variances to study the effect on the EP and expected Bernoulli variance.

@@

Table \ref{tab:sim_rhoab} shows the initial EPs and the associated
Bernoulli variance (second row) for the examples indicated in
Fig.~\ref{illus_bivarDens}. The EPs increase with the correlation as
there is a strong tendency to have jointly low or high temperature and
salinity. The Bernoulli variance is similarly larger for high
correlations. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, which implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

\begin{table}[!t] \centering \caption{EP and Bernoulli variance for different correlations and variances (top rows), and expected Bernoulli variances for both temperature and salinity data, and only temperature data (bottom rows).}
  \begin{tabular}{c|ccc|ccc}
 &\multicolumn{3}{c}{$\sigma_1=\sigma_2=1$} & \multicolumn{3}{c}{$\sigma_1=\sigma_2=2$} \\
\hline
Correlation $\gamma$ & 0.2 & 0.6 & 0.8 & 0.2 & 0.6 & 0.8 \\
\hline
$p$ & 0.28 & 0.35 & 0.40 & 0.28 & 0.35 & 0.40 \\ 
$p(1-p)$ & 0.20 & 0.23 & 0.24 & 0.20 & 0.23 & 0.24 \\ 
EBV, Temperature and Salinity & 0.092 & 0.089 & 0.085 & 0.052 & 0.051 & 0.049 \\ 
EBV, Temperature only & 0.151 & 0.138 & 0.123 & 0.137 & 0.114 & 0.093 \\ 
\hline
\end{tabular}
\label{tab:sim_rhoab}
\end{table}
The bottom two rows of Table \ref{tab:sim_rhoab} show 
expected Bernoulli variance results. This is presented for a design gathering
both data types, and for a design with temperature measurements
alone. When both data are gathered, the measurement model is
$(Y_1,Y_2)^t=(Z_1,Z_2)^t+\bepsilon$, with
$\bepsilon \sim N(0,0.5^2I_2)$, while $Y_1=Z_1+\epsilon$,
$\epsilon \sim N(0,0.5^2)$ when only temperature is measured.  For
this illustration, Table \ref{tab:sim_rhoab} shows that the expected
Bernoulli variance gets lower with larger standard deviations
$\sigma_1$ and $\sigma_2$. The expected reduction of Bernoulli variance is
largest for the cases with high correlation $\gamma$. Albeit smaller,
there is also uncertainty reduction when only temperature is measured
(bottom row), especially when temperature and salinity are highly
correlated. When correlation is low ($\gamma=0.2$), there is little
information about salinity in the temperature data, and therefore less
uncertainty reduction. 
%In an application with fresh cold water from a river source, the temperature and salinity variables will not only be interdependent, but will also likely show dependence in the spatial dimension. This in turn will impact the design criteria when we
% evaluate the information measure by integrating over several locations (Section \ref{sec:simulations}).

@@

We now turn to an example involving a full-fledged gaussian random
field (GRF). The class of GRF model we consider have a linear trend
\begin{align*}
\mu(s)=\mathbb{E}\left[\begin{pmatrix}
Z_{\spatloc, 1}\\ Z_{\spatloc, 2}
\end{pmatrix}\right] &= \beta_0 + \beta_1 \spatloc
\end{align*}
with $\beta_0$ a two dimensional vector and $\beta_1$ a $2\times 2$ matrix. In our examples, we only consider separable covariance models;
\begin{align*}
\textrm{Cov}\left(Z_{\spatloc, i}, Z_{v, j}\right) &= k(\spatloc, v) \gamma(i, j),~ \gamma(i, j) = \begin{cases} \sigma_i^2,~ i=j\\
   \gamma \sigma_i \sigma_j,~i\neq j,
        \end{cases}
\end{align*}
where $k(., .)$ is one of the traditional spatial covariance kernel
(exponential, Mat\'{e}rn, etc.) and $\gamma$ defines the
cross-covariance structure. 
In the accompanying Python
examples taking place within the MESLAS toolbox~\footnote{\url{https://github.com/CedricTravelletti/MESLAS}}, these
modeling assumptions can however be relaxed to anisotropic covariance and
changing variance levels across the spatial domain. Anisotropy and
non-stationary variance are both relevant for the setting with river
plumes, but in practice this requires more parameters to be
specified. With extensive satellite data or prior knowledge from high-resolution ocean models, one could also possibly fit more complex multivariate
spatial covariance functions
\citep{gneiting2010matern,genton2015cross}, but that is outside the
scope of the current paper.

In the rest of this section, we will consider a GRF with mean and
covariance structure as above and parameters
\begin{align*}
\beta_0 = \begin{pmatrix}
5.8\\ 24.0
\end{pmatrix}, ~ \beta_1 = \begin{pmatrix}
0.0 & -4.0\\
0.0 & -3.8
\end{pmatrix},~ \sigma_1 = \sigma_2 = 2.25, ~ \gamma_0 = 0.2
\end{align*}
and the spatial covariance has unit variance and a Mat\'{e}rn 3/2 kernel, $(1+\eta h)\exp (-\eta h)$, for Euclidean distance $h$ and decay parameter $\eta=0.5$.
One realization of this GRF is shown in Fig.~\ref{example_excu}.

We now study how the expected Bernoulli variance eq.\eqref{def:eibv}
associated with data collection at a point changes if only one of the
two components of the field is observed. We first draw a realization
of the GRF defined above and use it as ground-truth to mimic the real
data-collection process. A first set of observations are done at the
locations depicted in grey (see Fig.~\ref{fig:ebv_comp}), and the data
is used to update the GRF model. We then consider the green triangle
as a potential next observation location and plot the expected
Bernoulli variance reduction (at each grid point/nodes in the the waypoint graph) that would result from observing only one component of the field (temperature or salinity), or both at that point.

@@

Note that plotting the Bernoulli variance reduction at each point
might also be used to compare different data collection plans. For
example, Fig. \ref{fig:ebv_north_vs_east} shows the expected Bernoulli
variance reduction associated with a data collection plan along a
vertical line (static north) and one associated with a horizontal
(static east). Both expectations are computed according to the
a-priori distribution of the GRF (i.e. no observations have been
included yet).

\begin{figure}[!b] 
\centering 
\subfigure[Excursion probability.]{\includegraphics[
height=0.25\textwidth,keepaspectratio]{Figures/ebv_static_excu.png}\label{fig:ebv_static_excu}}
\subfigure[Static north design.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_north.png}\label{fig:ebv_static_north}}
\subfigure[Static east design.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_east.png}\label{fig:ebv_static_east}}
  \caption{Pointwise Bernoulli variance reduction for two different observation plans. The prior excursion probability is shown in \ref{fig:ebv_static_excu}. EBV reduction for each design shown in \ref{fig:ebv_static_north} and \ref{fig:ebv_static_east}.}
\label{fig:ebv_north_vs_east}
\end{figure}

>>>>

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime resource management. Faced with
  limited resources in understanding processes in the water-column,
  the combination of statistics and autonomous systems provide new
  opportunities for experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal ocean processes by providing informative descriptions of
  spatial characteristics of ocean phenomena. Specifically, we define
  a design criterion based on uncertainty in the excursions of
  vector-valued Gaussian random fields, and derive tractable
  expressions for the expected integrated Bernoulli variance reduction in such a
  framework. We demonstrate how this criterion can be used to
  prioritize sampling efforts at locations that are ambiguous, making
  exploration more effective. We use simulations to study the
  properties of methods and to compare them with state-of-the-art
  approaches, followed by results from field deployments with an
  autonomous underwater vehicle as part of a study mapping the
  boundary of a river plume. The results demonstrate the potential of
  combining statistical methods and robotic platforms to effectively
  inform and execute data-driven environmental sampling.

@@

\subsection{Expected Bernoulli variance for a two dimensional Example}
\label{Sec:UnivarEx}

We illustrate the expected Bernoulli variance (EBV) associated with different designs on a simple
bivariate example. This mimics our river plume application and hence
the first and second component of the random field will be called
\textit{temperature} and \textit{salinity}. We begin
with a \textit{pointwise} example, considering a single bivariate Gaussian distribution (i.e. no spatial elements).

@@

Say we want to study the excursion probability of a bivariate Gaussian
above some threshold, where the thresholds are set equal to the mean;
$\mu_1=t_1=5^o C$ for temperature and $\mu_2=t_2=30$ g/kg for
salinity, and we play with the temperature and salinity correlation
and variances to study the effect on the EP and EBV.

@@

Table \ref{tab:sim_rhoab} shows the initial EPs and the associated
Bernoulli variance (second row) for the examples indicated in
Fig.~\ref{illus_bivarDens}. The EPs increase with the correlation as
there is a strong tendency to have jointly low or high temperature and
salinity. The Bernoulli variance is similarly larger for high
correlations. EPs and Bernoulli variances are the same for temperature and salinity standard
deviations $\sigma_1$ and $\sigma_2$, which implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

\begin{table}[!t] \centering \caption{EP and Bernoulli variance for different correlations and variances (top rows), and EBVs for both temperature and salinity data, and only temperature data (bottom rows).}
  \begin{tabular}{c|ccc|ccc}
 &\multicolumn{3}{c}{$\sigma_1=\sigma_2=1$} & \multicolumn{3}{c}{$\sigma_1=\sigma_2=2$} \\
\hline
Correlation $\gamma$ & 0.2 & 0.6 & 0.8 & 0.2 & 0.6 & 0.8 \\
\hline
$p$ & 0.28 & 0.35 & 0.40 & 0.28 & 0.35 & 0.40 \\ 
$p(1-p)$ & 0.20 & 0.23 & 0.24 & 0.20 & 0.23 & 0.24 \\ 
EBV, Temperature and Salinity & 0.092 & 0.089 & 0.085 & 0.052 & 0.051 & 0.049 \\ 
EBV, Temperature only & 0.151 & 0.138 & 0.123 & 0.137 & 0.114 & 0.093 \\ 
\hline
\end{tabular}
\label{tab:sim_rhoab}
\end{table}
The bottom two rows of Table \ref{tab:sim_rhoab} show 
EBV results. This is presented for a design gathering
both data types, and for a design with temperature measurements
alone. When both data are gathered, the measurement model is
$(Y_1,Y_2)^t=(Z_1,Z_2)^t+\bepsilon$, with
$\bepsilon \sim N(0,0.5^2I_2)$, while $Y_1=Z_1+\epsilon$,
$\epsilon \sim N(0,0.5^2)$ when only temperature is measured.  For
this illustration, Table \ref{tab:sim_rhoab} shows that the expected
Bernoulli variance gets smaller with larger standard deviations. The expected reduction of Bernoulli variance is
further largest for the cases with high correlation $\gamma$. Albeit smaller,
there is also uncertainty reduction when only temperature is measured
(bottom row), especially when temperature and salinity are highly
correlated. When correlation is low ($\gamma=0.2$), there is little
information about salinity in the temperature data, and therefore less
uncertainty reduction. 
%In an application with fresh cold water from a river source, the temperature and salinity variables will not only be interdependent, but will also likely show dependence in the spatial dimension. This in turn will impact the design criteria when we
% evaluate the information measure by integrating over several locations (Section \ref{sec:simulations}).

@@

We now turn to an example involving a full-fledged GRF. We consider a linear trend
\begin{align*}
\mu(s)=\mathbb{E}\left[\begin{pmatrix}
Z_{\spatloc, 1}\\ Z_{\spatloc, 2}
\end{pmatrix}\right] &= \beta_0 + \beta_1 \spatloc,
\end{align*}
with $\beta_0$ a two dimensional vector and $\beta_1$ a $2\times 2$ matrix. In our examples, we only consider separable covariance models;
\begin{align*}
\textrm{Cov}\left(Z_{\spatloc, i}, Z_{v, j}\right) &= k(\spatloc, v) \gamma(i, j),~ \gamma(i, j) = \begin{cases} \sigma_i^2,~ i=j\\
   \gamma \sigma_i \sigma_j,~i\neq j,
        \end{cases}
\end{align*}
where an isotropic Mat\'{e}rn 3/2 kernel $(1+\eta h)\exp (-\eta h)$ is used, for Euclidean distance $h$.
In the accompanying Python
examples taking place within the MESLAS toolbox~\footnote{\url{https://github.com/CedricTravelletti/MESLAS}}, these
modeling assumptions can however be relaxed to anisotropic covariance and
changing variance levels across the spatial domain. Both extensions are relevant for the setting with river
plumes, but in practice this requires more parameters to be specified. With extensive satellite data or prior knowledge from high-resolution ocean models, one could also possibly fit more complex multivariate
spatial covariance functions
\citep{gneiting2010matern,genton2015cross}, but that is outside the
scope of the current paper.

In the rest of this section, we consider a GRF with mean and
covariance structure as above and parameters
\begin{align*}
\beta_0 = \begin{pmatrix}
5.8\\ 24.0
\end{pmatrix}, ~ \beta_1 = \begin{pmatrix}
0.0 & -4.0\\
0.0 & -3.8
\end{pmatrix},~ \sigma_1 = 2.5,~ \sigma_2 = 2.25, ~ \gamma = 0.2,
\end{align*}
and kernel parameter $\eta=3.5$.
One realization of this GRF is shown in Fig.~\ref{example_excu}.

We now study how the EBV [Eqn.\eqref{def:eibv}]
associated with data collection at a point changes if only one of the
two components of the field is observed. We first draw a realization
of the GRF defined above and use it as ground-truth to mimic the real
data-collection process. A first set of observations are done at the
locations depicted in grey (see Fig.~\ref{fig:ebv_comp}), and the data
is used to update the GRF model. We then consider the green triangle
as a potential next observation location and plot the EBV reduction (at each grid node in the waypoint graph) that would result from observing only one component of the field (temperature or salinity), or both at that point.

@@

Note that plotting the EBV reduction at each point
might also be used to compare different data collection plans. For
example, Fig. \ref{fig:ebv_north_vs_east} shows the EBV reduction associated with a data collection plan along a vertical line (static north) and one associated with a horizontal
(static east). Both expectations are computed according to the
a-priori distribution of the GRF (i.e. no observations have been
included yet).

\begin{figure}[!b] 
\centering 
\subfigure[Excursion probability.]{\includegraphics[
height=0.25\textwidth,keepaspectratio]{Figures/ebv_static_excu.png}\label{fig:ebv_static_excu}}
\subfigure[Static north design.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_north.png}\label{fig:ebv_static_north}}
\subfigure[Static east design.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_east.png}\label{fig:ebv_static_east}}
  \caption{Pointwise Bernoulli variance reduction for two different observation plans. The prior EP is shown in \ref{fig:ebv_static_excu}. EBV reduction for each design shown in \ref{fig:ebv_static_north} and \ref{fig:ebv_static_east}.}
\label{fig:ebv_north_vs_east}
\end{figure}
